---
title: "Foss_NIR_Data"
author: "Sydney Berry"
output: html_document
date: "2024-02-21"
---
# Load in Libraries
```{r libraries}
library(prospectr)
library(ggplot2)
library(tidyverse)
library(caret)
library(stringr)
```
# Load in Data
First I need to import the 2 .csv files into R.
```{r load in data}
raw_data <- read_csv("~/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/YCH_Scan_Data_part1.csv")
genotypes_data <- read_csv("~/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/YCH_Plot_2_Genotype.csv")
```
# Modify the Data
I need to modify and filter the data so I can work with it. This involves removing a row of the data that is not needed in the analysis and naming the first column.
```{r modify data}
#remove empty row
cleaned_data <- raw_data[-c(1), ]
#rename Wavelength row
cleaned_data <- cleaned_data %>%
  rename("Wavelength" = "...1")
#remove 'sample no:' from sample names
colnames(cleaned_data) <- gsub("sample no: ", "", colnames(cleaned_data))
```

# Pivot Table
I need to change the format of the current data so that ggplot() will be able to make a table with it. Also, I need to use mutate() to change the character types to numerical types. Finally, the filter() function will remove every wavelength that is not filled in on the dataset.
```{r pivot and modify data}
#create pivot table and filter wavelength for every 5th wavelength instead of every 0.5 nm
pivot_data <- pivot_longer(cleaned_data, cols = -c(Wavelength), names_to = 'SampleID', values_to = 'Absorbance') %>%
  mutate(Wavelength = as.numeric(Wavelength)) %>%
  mutate(Absorbance = as.numeric(Absorbance)) %>%
  filter(!is.na(Absorbance),
         Wavelength %% 5 == 0)
#add year to data frame by indexing sample name based on place of year number
pivot_data$Year <- ifelse(substr(pivot_data$SampleID, 3, 3) == "H",
                                substr(pivot_data$SampleID, 4, 5),
                                substr(pivot_data$SampleID, 3, 4))
#reorder data frame so year is 2nd column and remove NA years, "61", "69", and "mp"
pivot_data <- pivot_data[, c(2, 4, 1, 3)] %>%
  filter(!is.na(Year)) %>%
  subset(Year != "61" & Year != "69" & Year != "mp")
#removing any samples that aren't YCH and any sample names that include '...'
pivot_data <- pivot_data %>%
  filter(str_detect(SampleID, "YCH")) %>%
  mutate(SampleID = str_replace(SampleID, "\\.\\.\\.\\d+$", ""))
#add genotypes into pivot data
pivot_data <- left_join(pivot_data, genotypes_data, join_by(SampleID == Plot))
#remove any samples that have a genotype containing RIB
pivot_data <- pivot_data[!str_detect(pivot_data$Genotype, "RIB$"), ]
pivot_data <- pivot_data[!str_detect(pivot_data$Genotype, "FILL"), ]
#move genotypes column to front
pivot_data <- pivot_data[, c(1, 2, 5, 3, 4)]
#split genotypes into egg parent and pollen parent
pivot_data <- separate(pivot_data, Genotype, into = c("Egg_Parent", "Pollen_Parent"), sep = " [x,X] ", remove = FALSE)

#create wide data table
data_wide <- pivot_data %>%
  pivot_wider(id_cols = c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_from = Wavelength, values_from = Absorbance)
#unnest data table
data_wide <- unnest(data_wide, cols = everything())
#remove duplicate samples
data_wide <- data_wide %>%
  distinct(SampleID, .keep_all = TRUE)
#pivot back to long form
pivot_data <- pivot_longer(data_wide, cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_to = 'Wavelength', values_to = 'Absorbance') %>%
  mutate(Wavelength = as.numeric(Wavelength))
```

Now that the data has been modified and become a pivot chart, I can make a plot of the spectral data presented.
# Making the Spectral Plot
```{r spectral plot}
#plot unfiltered spectral plot
pivot_data %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Unfiltered Spectral Data")+
  facet_wrap(~Year)+
  theme_classic()
```

# PCA
Now I can use the data file from the last section to perform a PCA.
```{r pca}
#calculate PC's
unfiltered_data_pca <- prcomp(data_wide[,-c(1,2,3,4,5)], center = TRUE, scale. = TRUE)
```
# PCA Plot
```{r pca plot}
#plot PC1 vs PC2
data_wide %>%
  select(1:5) %>%
  bind_cols(as_tibble(unfiltered_data_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, col = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Unfilitered Spectral Data') +
  theme_classic()
```

# Filtering Outliers
I will now remove any outlier spectra from the NIR data so that I know those are not included in the sample selection. I will do this using the Mahalanobis Distance Code.
```{r mahalanobis distance for outlier removal}
#waveband matrix
wavebands <- (seq(400, 1099.5, 5))
#calculate the mahalanobis distance
m_dist <- as.matrix(mahalanobis(data_wide[,colnames(data_wide) %in% wavebands],                                colMeans(data_wide[,colnames(data_wide) %in% wavebands]),
                    cov(data_wide[,colnames(data_wide) %in% wavebands]), na.rm=TRUE))
#appends Mdist to data_wide
data_wide$Mdist = round(m_dist, 1)
#create data frame without outliers
outliers_removed_data <- data_wide %>%
  mutate(MThresh = 3 * mean(Mdist)) %>%
  filter(Mdist < MThresh) %>%
  select(-Mdist,
         -MThresh) #this removed 27 samples
```
# Verify Removal of Outliers
Now that the outliers are removed, I will make sure that these are removed from the sample set by looking at the spectral plot and PCA again.
```{r pivot clean data}
#create pivot table for outliers removed
outliers_removed_pivot <- pivot_longer(outliers_removed_data, cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_to = 'Wavelength', values_to = 'Absorbance') %>%
  filter(!is.na(Absorbance)) %>%
  mutate(Wavelength = as.numeric(Wavelength))

#make spectral plot
outliers_removed_pivot %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Cleaned Spectral Data")+
  scale_x_continuous(breaks=seq(400, 1095, 100))+
  scale_y_continuous(breaks=seq(0, 10, 1))+
  facet_wrap(~Year)+
  theme_classic()
```

```{r clean pca} 
#calculate PC's for outliers removed data
pca_clean <- prcomp(outliers_removed_data[,-c(1:5)], center = TRUE, scale. = TRUE)
#plot PC1 vs PC2
outliers_removed_data %>%
  select(1:5) %>%
  bind_cols(as_tibble(pca_clean$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Cleaned Spectral Data') +
  theme_classic()
```

# Normalizing the Data
There are multiple different normalization features in the prospectr package. I'm going to test all of these and determine which one works the best for this data.
Baseline, Detrend, SNV
```{r baseline spectral}
#Estimates the baseline of a given spectrum and subtracts it from the original spectrum. Pulls convex hull points of each spectrum are identified, these are linearly interpolated to the same frequencies of the original spectra. Linearly interpolated: means that intermediate data between known values is connected with a conceptual straight line.Baseline of each spectrum is subtracted from the original inpur spectrum. Aims to reset all of the spectra on a common baseline. Caution in interpreting because it can distort the real proportions between absorbance peaks.
#Normalizations
b_normalization <- baseline(outliers_removed_data[,-c(1:5)], wav = as.numeric(colnames(outliers_removed_data[,-c(1:5)])))

#Combine data sets to include Sample ID
baseline_normalization <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(b_normalization))

#Pivot Table
baseline_normalization_pivot <- baseline_normalization %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))

#Spectral plot
baseline_normalization_pivot %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Baseline Normalized Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r baseline pca}
#Check for zero variance
NZV_baseline <- nearZeroVar(baseline_normalization)
baseline_normalization <- baseline_normalization[, -NZV_baseline]
#PCA
baseline_pca = prcomp(baseline_normalization[,-c(1:5)], scale. = TRUE, center = TRUE)
#Plot PCA
baseline_normalization %>%
  select(1:5) %>%
  bind_cols(as_tibble(baseline_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Baseline-Normalized Data') +
  theme_classic()
```

```{r detrend spectra}
#Accounts for wavelength-dependent scattering effects. Corrects for curvilinear effects in spectra of tightly packed samples. Curvature of spectra depends on packing density and particle sizes of a specific samples, detrend standardizes the variation in the curvilinearity. Aims to show only the differences in values from the trend.
#Normalization
d_normalization <- detrend(outliers_removed_data[,-c(1:5)], wav = as.numeric(colnames(outliers_removed_data[,-c(1:5)], p = 2)))
#Combine data sets to include SampleID
detrend_normalization <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(d_normalization))
#Pivot table
detrend_normalization_pivot <- detrend_normalization %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))
#Spectral plot
detrend_normalization_pivot %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Detrend Normalized Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r detrend pca}
#PCA
detrend_pca = prcomp(detrend_normalization[,6:145], scale. = TRUE, center = TRUE)
#Plot PCA
detrend_normalization %>%
  select(1:5) %>%
  bind_cols(as_tibble(detrend_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Detrend-Normalized Data') +
  theme_classic()
```

```{r snv spectral}
#SNV tries to correct for light scatter on a row-wise operation. Removes the multiplicative interferences of scatter and particle size. Compared to original datasets, there is no divergence between similar samples with varying particle size, but the overall shape does not change.
#Normalization
s_normalization <- standardNormalVariate(outliers_removed_data[,-c(1:5)])
#Combine data sets to include SampleID
snv_normalization <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(s_normalization))
#Pivot table
snv_normalization_pivot <- snv_normalization %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))
#Spectral plot
snv_normalization_pivot %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "SNV Normalized Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r snv pca}
#PCA
snv_pca = prcomp(snv_normalization[,6:145], scale. = TRUE, center = TRUE)
#Plot PCA
snv_normalization %>%
  select(1:5) %>%
  bind_cols(as_tibble(snv_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of SNV-Normalized Data') +
  theme_classic()
```

```{r baseline on snv spectral}
#Normalizations
baseline_on_snv <- baseline(s_normalization, wav = as.numeric(colnames(s_normalization)))
#Combine data sets to include Sample ID
baseline_on_snv <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(baseline_on_snv))
#Pivot Table
baseline_on_snv <- baseline_on_snv %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))
#Spectral plot
baseline_on_snv %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Baseline on SNV Normalized Data Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r baseline on snv pca}
#Pivot to wide format
baseline_on_snv_wide <- baseline_on_snv %>%
  pivot_wider(id_cols = c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_from = Wavelength, values_from = Absorbance)
baseline_on_snv_wide <- unnest(baseline_on_snv_wide, cols = everything())
#Check for zero variance
NZV_baseline_snv <- nearZeroVar(baseline_on_snv_wide)
baseline_on_snv_wide <- baseline_on_snv_wide[, -NZV_baseline_snv]
#PCA
baseline_snv_pca = prcomp(baseline_on_snv_wide[,6:134], scale. = TRUE, center = TRUE)
#Plot PCA
baseline_on_snv_wide %>%
  select(1:5) %>%
  bind_cols(as_tibble(baseline_snv_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Baseline on SNV Normalized Data') +
  theme_classic()
```

```{r snv on baseline spectral}
#Normalization
snv_on_baseline <- standardNormalVariate(b_normalization)
#Combine data sets to include Sample ID
snv_on_baseline <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(snv_on_baseline))
#Pivot Table
snv_on_baseline <- snv_on_baseline %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))
#Spectral plot
snv_on_baseline %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "SNV on Baseline Normalized Data Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r snv on baseline pca}
#Pivot to wide format
snv_on_baseline_wide <- snv_on_baseline %>%
  pivot_wider(id_cols = c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_from = Wavelength, values_from = Absorbance)
snv_on_baseline_wide <- unnest(snv_on_baseline_wide, cols = everything())
#PCA
snv_baseline_pca = prcomp(snv_on_baseline_wide[,6:145], scale. = TRUE, center = TRUE)
#Plot PCA
snv_on_baseline_wide %>%
  select(1:5) %>%
  bind_cols(as_tibble(snv_baseline_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of SNV on Baseline Normalized Data') +
  theme_classic()
```

```{r baseline on detrend spectral}
#Normalization
baseline_on_detrend <- baseline(d_normalization, wav = as.numeric(colnames(d_normalization)))
#Combine data sets to include Sample ID
baseline_on_detrend <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(baseline_on_detrend))
#Pivot Table
baseline_on_detrend <- baseline_on_detrend %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))
#Spectral plot
baseline_on_detrend %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Baseline on Detrend Normalized Data Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r baseline on detrend pca}
#Pivot to wide format
baseline_on_detrend_wide <- baseline_on_detrend %>%
  pivot_wider(id_cols = c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_from = Wavelength, values_from = Absorbance)
baseline_on_detrend_wide <- unnest(baseline_on_detrend_wide, cols = everything())
#Check for zero variance
NZV_baseline_detrend <- nearZeroVar(baseline_on_detrend_wide)
baseline_on_detrend_wide <- baseline_on_detrend_wide[, -NZV_baseline_detrend]
#PCA
baseline_detrend_pca = prcomp(baseline_on_detrend_wide[,6:143], scale. = TRUE, center = TRUE)
#Plot PCA
baseline_on_detrend_wide %>%
  select(1:5) %>%
  bind_cols(as_tibble(baseline_detrend_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Baseline on Detrend Normalized Data') +
  theme_classic()
```

```{r detrend on baseline spectral}
#Normalization
detrend_on_baseline <- detrend(b_normalization, wav = as.numeric(colnames(b_normalization)))
#Combine data sets to include SampleID
detrend_on_baseline <- outliers_removed_data %>%
  select("SampleID", "Year", "Genotype", "Egg_Parent", "Pollen_Parent") %>%
  bind_cols(as_tibble(detrend_on_baseline))
#Pivot table
detrend_on_baseline <- detrend_on_baseline %>%
  pivot_longer(cols = -c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent),
               names_to = "Wavelength",
               values_to = "Absorbance") %>%
  mutate(Wavelength = as.numeric(Wavelength))
#Spectral plot
detrend_on_baseline %>%
  ggplot(aes(x = Wavelength, y = Absorbance, group = SampleID))+
  geom_line(aes(col=Year))+
  labs(title = "Detrend on Baseline Normalized Data Spectral Plot")+
  facet_wrap(~Year)+
  theme_classic()
```

```{r detrend on baseline pca}
#Pivot to wide format
detrend_baseline_wide <- detrend_on_baseline %>%
  pivot_wider(id_cols = c(SampleID, Year, Genotype, Egg_Parent, Pollen_Parent), names_from = Wavelength, values_from = Absorbance)
detrend_baseline_wide <- unnest(detrend_baseline_wide, cols = everything())
#PCA
detrend_baseline_pca = prcomp(detrend_baseline_wide[,6:145], scale. = TRUE, center = TRUE)
#Plot PCA
detrend_baseline_wide %>%
  select(1:5) %>%
  bind_cols(as_tibble(detrend_baseline_pca$x)) %>%
  ggplot(aes(x = PC1, y = PC2, color = Year)) +
  geom_point() +
  stat_ellipse() +
  labs(title = 'PCA of Detrend on Baseline Normalized Data') +
  theme_classic()
```

Now that the data is normalized, I can look at which may be the best option going forward. Due to the PCA's of the normalizations being pretty well overlapped by year, there is not one option that seems to be better than the rest. I will attempt to pull spectrally diverse samples out of each dataset and see how normalizations affect the data.

# Honigs Regression
Honigs pulls samples based on the size of their absorption features. It selects the sample with the highest absorption feature first, subtracts it from the other spectra and the function continues to select samples with the highest absorption
```{r honigs on outliers removed data}
#add row number for honigs
outliers_removed_data <- outliers_removed_data %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
outliers_removed_data <- outliers_removed_data[, c(1:5, ncol(outliers_removed_data), 6:(ncol(outliers_removed_data) - 1))]
#honigs regression
honigs_outliers = honigs(outliers_removed_data[,-c(1:6)], k = 60, type = "A")
honigs_outliers_removed = outliers_removed_data %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_outliers$model ~ 'Train',
                           row_num %in% honigs_outliers$test ~ 'Test'),
         .before = '400')
#create sample selection data table
sample_selection_outliers <- honigs_outliers_removed %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot honigs regression spectral plot
honigs_outliers_removed %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband),
         Group = factor(Group, levels = c('Test', 'Train'))) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num, alpha = Group)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  scale_alpha_manual(values = c(1, 0.3), breaks = c('Train', 'Test'))+
  labs(title = "Selected Samples from Outliers Dataset") +
  facet_wrap(~Group) +
  theme_classic()
```

```{r pca of honigs outliers}
#pca for honigs outliers
pca_honigs_outliers <- prcomp(honigs_outliers_removed[,8:147], center = T, scale. = T)
#plot PC1 vs PC2
honigs_outliers_removed %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_outliers$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from Outliers Removed Data") +
  theme_classic()
```

```{r honigs on baseline}
#row number addition
baseline_normalization <- baseline_normalization %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
baseline_normalization <- baseline_normalization[, c(1:5, ncol(baseline_normalization), 6:(ncol(baseline_normalization) - 1))]
#honigs
honigs_b = honigs(baseline_normalization[,-c(1:6)], k = 60, type = "A")
honigs_baseline = baseline_normalization %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_b$model ~ 'Train',
                           row_num %in% honigs_b$test ~ 'Test'),
         .before = '405') %>%
  mutate(row_num = as.character(row_num))
#create sample selection data table
sample_selection_baseline <- honigs_baseline %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot honigs baseline
honigs_baseline %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from Baseline Dataset") +
  facet_wrap(~Group)+
  theme_classic()
```

```{r honigs baseline pca}
#pc calculation
pca_honigs_baseline <- prcomp(honigs_baseline[,8:136], center = T, scale. = T)
#plot PC1 vs PC2
honigs_baseline %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_outliers$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from Baseline Data") +
  theme_classic()
```

```{r honigs detrend}
#add row number column
detrend_normalization <- detrend_normalization %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
detrend_normalization <- detrend_normalization[, c(1:5, ncol(detrend_normalization), 6:(ncol(detrend_normalization) - 1))]
#honigs regression
honigs_d = honigs(detrend_normalization[,-c(1:6)], k = 60, type = "A")
honigs_detrend = detrend_normalization %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_d$model ~ 'Train',
                           row_num %in% honigs_d$test ~ 'Test'),
         .before = '400') %>%
  mutate(row_num = as.character(row_num))
#sample selection
sample_selection_detrend <- honigs_detrend %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot spectral
honigs_detrend %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from Detrend Dataset") +
  facet_wrap(~Group)+
  theme_classic()
```

```{r pca honigs detrend}
#pca calculation
pca_honigs_detrend <- prcomp(honigs_detrend[,8:147], center = T, scale. = T)
#plot PC1 vs PC2
honigs_detrend %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_detrend$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from Detrend Data") +
  theme_classic()
```

```{r honigs snv}
#row number addition
snv_normalization <- snv_normalization %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
snv_normalization <- snv_normalization[, c(1:5, ncol(snv_normalization), 6:(ncol(snv_normalization) - 1))]
#honigs regression, pulling 75 samples because we are using this for sample selection
honigs_s = honigs(snv_normalization[,-c(1:6)], k = 60, type = "A")
honigs_snv = snv_normalization %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_s$model ~ 'Train',
                           row_num %in% honigs_s$test ~ 'Test'),
         .before = '400') %>%
  mutate(row_num = as.character(row_num))
#create data table
sample_selection_snv <- honigs_snv %>%
  slice(honigs_s$model) %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot spectral
honigs_snv %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(desc(Group)) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from SNV Dataset") +
  facet_wrap(~Group)+
  theme_classic()
```

```{r pca on snv}
#calculate PC
pca_honigs_snv <- prcomp(honigs_snv[,8:147], center = T, scale. = T)
#PC1 vs PC2
honigs_snv %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_snv$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from SNV Data") +
  theme_classic()
```

#SNV seems to give the best distribution of samples within the PC, so we will be using that for sample selection, but will calculate the other options as well.
```{r baseline on snv honigs}
#add row number
baseline_on_snv_wide <- baseline_on_snv_wide %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
baseline_on_snv_wide <- baseline_on_snv_wide[, c(1:5, ncol(baseline_on_snv_wide), 6:(ncol(baseline_on_snv_wide) - 1))]
#honigs regression
honigs_b_s = honigs(baseline_on_snv_wide[,-c(1:6)], k = 60, type = "A")
honigs_baseline_snv = baseline_on_snv_wide %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_b_s$model ~ 'Train',
                           row_num %in% honigs_b_s$test ~ 'Test'),
         .before = '405') %>%
  mutate(row_num = as.character(row_num))
#sample selection data table
sample_selection_baseline_snv <- honigs_baseline_snv %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot honigs
honigs_baseline_snv %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from Baseline on SNV Dataset") +
  facet_wrap(~Group)+
  theme_classic()
```

```{r honigs pca on baseline-snv}
#pca calculation
pca_honigs_baseline_snv <- prcomp(honigs_baseline_snv[,8:136], center = T, scale. = T)
#plot PC1 vs PC2
honigs_baseline_snv %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_baseline_snv$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from Baseline on SNV Data") +
  theme_classic()
```

```{r snv on baseline honigs}
#add row number
snv_on_baseline_wide <- snv_on_baseline_wide %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
snv_on_baseline_wide <- snv_on_baseline_wide[, c(1:5, ncol(snv_on_baseline_wide), 6:(ncol(snv_on_baseline_wide) - 1))]
#honigs regression
honigs_s_b = honigs(snv_on_baseline_wide[,-c(1:6)], k = 60, type = "A")
honigs_snv_baseline = snv_on_baseline_wide %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_s_b$model ~ 'Train',
                           row_num %in% honigs_s_b$test ~ 'Test'),
         .before = '400') %>%
  mutate(row_num = as.character(row_num))
#sample selection data table
sample_selection_snv_baseline <- honigs_snv_baseline %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot spectral data
honigs_snv_baseline %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from SNV on Baseline Dataset") +
  facet_wrap(~Group)+
  theme_classic()
```

```{r pca snv on baseline}
#pca calculation
pca_honigs_snv_baseline <- prcomp(honigs_snv_baseline[,8:147], center = T, scale. = T)
#plot PC1 vs PC2
honigs_snv_baseline %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_snv_baseline$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from SNV on Baseline Data") +
  theme_classic()
```

```{r honigs baseline on detrend}
#add row number
baseline_on_detrend_wide <- baseline_on_detrend_wide %>%
  ungroup() %>%
  mutate(row_num = row_number())
#reorder data table
baseline_on_detrend_wide <- baseline_on_detrend_wide[, c(1:5, ncol(baseline_on_detrend_wide), 6:(ncol(baseline_on_detrend_wide) - 1))]
#honigs regression
honigs_b_d = honigs(baseline_on_detrend_wide[,-c(1:6)], k = 60, type = "A")
honigs_baseline_on_detrend = baseline_on_detrend_wide %>%
  mutate(row_num = row_number(),
         Group = case_when(row_num %in% honigs_b_d$model ~ 'Train',
                           row_num %in% honigs_b_d$test ~ 'Test'),
         .before = '405') %>%
  mutate(row_num = as.character(row_num))
#sample selection data table
sample_selection_baseline_detrend <- honigs_baseline_on_detrend %>%
  filter(Group == 'Train') %>%
  select(SampleID, Genotype)
#plot spectral
honigs_baseline_on_detrend %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from Baseline on Detrend Dataset") +
  facet_wrap(~Group)+
  theme_classic()
```

```{r baseline on detrend pca honigs}
#calculate PC
pca_honigs_baseline_detrend <- prcomp(honigs_baseline_on_detrend[,8:145], center = T, scale. = T)
#plot PC1 vs PC2
honigs_baseline_on_detrend %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_baseline_detrend$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Selected Samples from Baseline on Detrend Data") +
  theme_classic()
```

#Now that we have calculated all of the honigs regression and come up with a list of samples for cooking, I'd like to compare all of the sample selection data tables together to see if there is any overlap between the different normalization methods.
```{r honigs upset plot}
#packages for UpSetR
library(UpSetR)
library(data.table)
#create big data table containing all sample selections
selected_samples_list <- data.table(
  sample_selection_baseline,
  sample_selection_baseline_detrend,
  sample_selection_baseline_snv,
  sample_selection_detrend,
  sample_selection_outliers,
  sample_selection_snv,
  sample_selection_snv_baseline
)
#change column names
colnames(selected_samples_list) <- c("Baseline", "Genotype_Baseline", "Baseline-Detrend", "Genotype_Baseline-Detrend", "Baseline-SNV", "Genotype_Baseline-SNV", "Detrend", "Genotype_Detrend", "Outliers", "Genotype_Outliers", "SNV", "Genotype_SNV", "SNV-Baseline", "Genotype_SNV-Baseline")
#create list to pull from
fld <- fromList(as.list(selected_samples_list))
upset(fld, nsets=14, matrix.color = "lightblue")
```

There is not a single sample or genotype that is the exact same between all of the samples selected and the genotypes correlated to them. This means that each group was very different from each other. We will move forward with SNV-Normalization. Upon examining the sample selection, we found that there was overlap of genotypes, specifically the egg parent. To fix this, we will be using a 'for loop' to remove all samples pertaining to an egg parent once it has been selected, so that all samples selected are spectrally diverse and have different egg parents.

```{r honigs snv removing genotypes}
#Modifying copy of snv_normalization data frame
copied_snv_normalization <- snv_normalization[snv_normalization$Pollen_Parent %in% c('B73', 'Mo17', 'LH244'),]
#copied_snv_normalization_sub = copied_snv_normalization
#this needs to be run with each modification of the loop
sample_selection_good = tibble()
#for loop to select spectrally diverse samples that do not have the same egg parent as any other sample selected
for(year in c('22', '23')){
  pollen_parents = unique(copied_snv_normalization$Pollen_Parent[copied_snv_normalization$Year == year])
  
  for(p in pollen_parents){

    for (i in 1:20) {
      print(paste('---', i, '---'))
      copied_snv_normalization_sub <- copied_snv_normalization[copied_snv_normalization$Year == year &
                                                                 copied_snv_normalization$Pollen_Parent == p,]

      
      result <- honigs(copied_snv_normalization_sub[,-c(1:6)], k = i+1, type = "A")
      
      #print(copied_snv_normalization_sub[result$model[1:i],]$SampleID)
      
      egg_parent <- copied_snv_normalization_sub[result$model[i], c(1,4,5)]
      
      sample_selection_good <- bind_rows(sample_selection_good, egg_parent)
      
      copied_snv_normalization <- copied_snv_normalization[copied_snv_normalization$Egg_Parent != egg_parent$Egg_Parent | copied_snv_normalization$SampleID == egg_parent$SampleID,]
    }
  }
}

length(unique(sample_selection_good$SampleID))

#make new file for spectral plot
honigs_snv_good = snv_normalization %>%
  mutate(Group = case_when(SampleID %in% sample_selection_good$SampleID ~ 'Train',
                           !SampleID %in% sample_selection_good$SampleID ~ 'Test'),
         .before = '400') %>%
  mutate(row_num = as.character(row_num))

honigs_snv_good %>%
  filter(Group == 'Train') %>%
  group_by(Year) %>%
  summarise(n = n())

honigs_snv_good %>%
  filter(Group == 'Train') %>%
  group_by(Pollen_Parent) %>%
  summarise(n = n())

honigs_snv_good %>%
  filter(Group == 'Train') %>%
  group_by(Year, Pollen_Parent) %>%
  summarise(n = n())
```

```{r plot new honigs snv}
honigs_snv_good %>%
  pivot_longer(cols = -(1:7), names_to = 'Waveband', values_to = 'Absorbance') %>%
  mutate(Waveband = as.numeric(Waveband)) %>%
  arrange(Group) %>%
  ggplot(aes(x = Waveband, y = Absorbance, color = Group, group = row_num)) +
  geom_line() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "Selected Samples from Final SNV Sample Selection") +
  facet_wrap(~Group) +
  theme_classic()
```

```{r pca for final snv sample selection}
#calculate PC
pca_honigs_snv_final <- prcomp(honigs_snv_good[,8:147], center = T, scale. = T)
#plot PC1 vs PC2
honigs_snv_good %>%
  select(1:7) %>%
  bind_cols(as_tibble(pca_honigs_snv_final$x)) %>%
  arrange(Group) %>%
  ggplot(aes(x = PC1, y = PC2, color = Group, shape = Year)) +
  geom_point() +
  scale_color_manual(values = c('red', 'black'), breaks = c('Train', 'Test')) +
  labs(title = "PCA of Final Sample Selection") +
  theme_classic()
```

Now that we have successfully identified all of the samples we will use for the cook validation, I will export the final sample selection as a .csv file.
```{r export csv file}
#add full genotype information
sample_selection_good <- left_join(sample_selection_good, genotypes_data, join_by(SampleID == Plot))
#export as .csv file
sample_selection_good %>%
  write_csv("~/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/methodology_validation_samples2find.csv")
```

```{r read in found validation samples}
samples = read_csv('~/Downloads/Pericarp Methods Validation Samples to Find - Sheet1.csv') %>%
  filter(Found == 1) %>%
  select(SampleID)
```

```{r create benchtop cook test labels}
set.seed(1234)
# Make labels for cook test packaging
sample_labels = samples %>%
   mutate(A = NA,
          B = NA) %>%
   pivot_longer(cols = c(A, B),
                names_to = 'Hotplate_ID') %>%
   select(-value) %>%
   sample_frac(1) %>%
   group_by(Hotplate_ID) %>%
   mutate(Hotplate_Pos = rep_len(c(1:4), n())) %>%
   group_by(Hotplate_ID, Hotplate_Pos) %>%
   mutate(Cook_Day = row_number()) %>%
   arrange(Cook_Day, Hotplate_ID, Hotplate_Pos)

# Write labels for packaging
# sample_labels %>%
#     write_csv('/Users/michael/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/Pericarp_Validation_Samples.csv')

# Create a write labels for subsamples
sample_labels %>%
   mutate(Pericarp_1 = NA,
          Pericarp_2 = NA,
          Kernels_1 = NA,
          Kernels_2 = NA) %>%
   pivot_longer(cols = c(Pericarp_1, Pericarp_2, Kernels_1, Kernels_2),
                names_to = 'Subsample') %>%
   select(-value) #%>%
   #write_csv('/Users/michael/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/Pericarp_Validation_Subsamples.csv')
```

```{r initial pericarp labels}
set.seed(12345)
initial_value_samples = samples %>%
  mutate(A = NA,
         B = NA) %>%
  pivot_longer(cols = c(A, B),
              names_to = 'Replicate') %>%
  select(-value) %>%
  sample_frac(1) %>%
  mutate(order = row_number()) %>%
  arrange(SampleID)

initial_value_samples #%>%
  #write_csv('/Users/michael/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/Pericarp_Validation_Initial_Pericarp_Labels.csv')


initial_value_samples %>%
   mutate(Pericarp_1 = NA,
          Pericarp_2 = NA) %>%
   pivot_longer(cols = c(Pericarp_1, Pericarp_2),
                names_to = 'Subsample') %>%
   select(-value) %>%
  arrange(order, Subsample)# %>%
  # write_csv('/Users/michael/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/Pericarp_Validation_Initial_Pericarp_Subsample_Labels.csv')
```

```{r rapid cook test labels}
set.seed(123456)
rapid_samples = samples %>%
  mutate(A = NA,
         B = NA) %>%
  pivot_longer(cols = c(A, B),
              names_to = 'Replicate') %>%
  select(-value) %>%
  sample_frac(1) %>%
  mutate(day = rep_len(c(1:4), n())) %>%
  arrange(SampleID)

rapid_samples %>%
  write_csv('/Users/michael/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/Pericarp_Validation_Rapid_Pericarp_Labels.csv')


rapid_samples %>%
   mutate(Pericarp_1 = NA,
          Pericarp_2 = NA,
          Kernels_1 = NA,
          Kernels_2 = NA) %>%
   pivot_longer(cols = c(Pericarp_1, Pericarp_2, Kernels_1, Kernels_2),
                names_to = 'Subsample') %>%
   select(-value) %>%
  arrange(day, Replicate) %>%
  write_csv('/Users/michael/Desktop/Grad_School/Research/Pericarp/Chemometrics/Data/Pericarp_Validation_Rapid_Pericarp_Subsample_Labels.csv')
```

